Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-03-18 analysis/paper/paper.Rmd
citations added

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287]. Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a long time it was believed the town contracted, moved, or even dissolved following the collapse of the palace. Indeed, Alexiou (1950 CITATION NEEDED) thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.

@@

The 134 tombs  that make up these two cemeteries, some of which were in use for several centuries, suggest an established, relatively stable community for whom tradition formed an important foundation in mortuary beliefs and behaviours [@Coldstream1994, 106-108]. The tombs fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. The chamber tomb comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or a combination of both (Antoniadis 2012, 47–8). It is not uncommon for dromoi to have niches cut into them, to house later burials that could not fit in the chamber (e.g. T. II), or even, in some cases, cinerary urns removed from the chamber so that new interments could be fitted in (e.g. T. TFT; Brock 1957, 3–4).

The majority of burials in these two cemeteries are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts. 200, 201 and 202 are three chambers of a single pit tomb) in the KNC (Coldstream 1994, 109). Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials, and in those instances where it is attested (Ts. L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb (Antoniadis 2012, 69). This degree of coexistence suggests that the shift between the two rites need not have marked a major ideological break; the preference for communal chamber tombs, for instance, something practiced since the Middle Minoan period at Knossos, remained constant. Whether this is linked to a strong sense of historical continuity or tradition, close kinship ties, or simply the need to bury large numbers of people in a small geographical area, we cannot be sure.

@@

The chronological sequence of the cemeteries has received much attention. It has been claimed that no tomb crosses the LM IIIC-SM divide (Coldstream 1998, 58; Coldstream 2006, 582), but there is a strong possibility that this is a classificatory artefact (Hallager 2010). The earliest tomb appears to be the T. 200+ complex, with its precocious cremations and imported artefacts. There remain ongoing debates about whether the KNC represents a foundation on virgin soil, as instances of BA tombs reused the EIA are known (e.g. the Khaniale Teke tholos). Cavanagh (1996), for one, could not rule out the possibility of some KNC tombs being of BA date based on his cluster analysis, but other evidence would suggest this is unlikely.

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles, and patterns of tomb construction and abandonment. In the Protogeometric period, some 22 tombs then in use in the KNC, and nine at Fortetsa, ceased to receive interments, followed in the succeeding Protogeometric B period by the founding of 22 new tombs (16 and six respectively). The PGB was defined by Brock (1957, 143), who hailed it as “the most remarkable phase in Cretan vase-painting”. Apparently a Cretan anomaly, this phase sees Attic geometric motifs mixed with experimental patterns possibly deriving from Near Eastern metalwork, as well as iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4). Coldstream has argued that the concurrence of this style and a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T. 107; see Coldstream 1994, 112–3; 1984). Whether simply nostalgia (Coldstream 1998, 60), or a more tactical strategy by newly emerging elite groups (1994, 114–5), it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.

The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of new burials around 630 BCE. There had been changes in tomb use in both cemeteries in the preceding century, with no new tombs built in Fortetsa, and few in the KNC, though interments in old tombs continued (Cavanagh 1996, 651–3; Brock 1957, 4). This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap’ (Coldstream, Huxley, and Webb 1999). The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown. Although various explanations have been forwarded, (Huxley 1994, 126; Coldstream, Huxley, and Webb 1999, 301–2; Kotsonas 2002, 41–44), it seems there are now strong reasons to doubt the settlement was truly abandoned (Whitelaw, pers comm.).

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai (1986) and Kopytoff (1986). Influenced by Mauss’ (1954 [1925]) formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence. In Aegean archaeology, these ideas proved attractive because of their resonance with the Homeric motif of gift-giving accompanied by accounts of an object's illustrious heritage (Whitley 2002, 220–1). It is hardly suprising that, when direct parallels for such objects have appeared archaeologically – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (Il. 10.260-271) – a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.

Catling’s discussion of the T. 200+ complex is a notable example. Tomb 201 contained the remains of at least two adults, and possibly a child. The burial objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned BA-style boar’s tusk helmet (Catling 1995, 123). Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi. All contained imported items at an early date, were of ‘warrior grave’ type, and formed loci for later burials. Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing with precious artefacts and stories of distant lands – ‘heroes returned’ in Catling’s phrase – would have been accorded exceptional treatment in death (Catling 1995, 127–8).

Complementing the theme of heroes abroad has been a similar degree of interest in foreigners at home. Two cache of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman (1967) to be foundation deposits – a known Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician, practising his trade at Knossos. Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output (Kotsonas 2006, 155–9). While certainty in such individual cases may be moot, it seems unlikely that traders or craftsmen of non-Cretan heritage never visited or resided at Knossos during the EIA (Hoffman 1997, 176–85; Schreiber 2003, 293-306).

Finally, a recurrent interest has been in the 'presence of the past' in these tombs. The practice of hero or ancestor cult, much discussed on the mainland (see Antonaccio 1994; Coldstream 1976), is not common here. Instead, we see several Late Minoan larnakes (and a few other vessels) redeposited in tombs and pits in the KNC and Fortetsa from the PGB period onward. A thorough analysis of these artefacts has been provided by Crowe (2016), who argues that the larnakes were robbed from tombs in the Knossos area, perhaps with a particular preference for painted specimens where available. A particularly evocative case comres from tomb 107, where parts of an elaborate figured LM larnax have been found along with a PGB vessel by the so-called Tree Painter. Both feature women in the company of birds and spiralling trees, and it seems certain that the former inspired the design of the latter. The PGB vessel also seems to draw on Oriental motifs, with the polos crown worn by the goddess having Near Eastern origins (Coldstream 1984a, 94–101). The magpie-like amalgamation of these styles and motifs seem to exemplify the particular cultural openness, fluidity and competition suggested to characterise the PGB period. Artefacts such as this hint towards an intimate discourse of (re)discovery; a painter, struck by the strangeness, or perhaps the uncanny familiarity, of centuries-old images, incorporated them into his own work.

>>>>

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287]. Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a long time it was believed the town contracted, moved, or even dissolved following the collapse of the palace. Indeed, Alexiou [@1950] thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.

@@

The 134 tombs  that make up these two cemeteries, some of which were in use for several centuries, suggest an established, relatively stable community for whom tradition formed an important foundation in mortuary beliefs and behaviours [@Coldstream1994, 106-108]. The tombs fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. The chamber tomb comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or a combination of both [@Antoniadis2012, 47-8]. It is not uncommon for dromoi to have niches cut into them, to house later burials that could not fit in the chamber (e.g. T. II), or even, in some cases, cinerary urns removed from the chamber so that new interments could be fitted in (e.g. T. TFT; @Brock1957, 3-4).

The majority of burials in these two cemeteries are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts. 200, 201 and 202 are three chambers of a single pit tomb) in the KNC [@Coldstream1994, 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials, and in those instances where it is attested (Ts. L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb [@Antoniadis2012, 69]. This degree of coexistence suggests that the shift between the two rites need not have marked a major ideological break; the preference for communal chamber tombs, for instance, something practiced since the Middle Minoan period at Knossos, remained constant. Whether this is linked to a strong sense of historical continuity or tradition, close kinship ties, or simply the need to bury large numbers of people in a small geographical area, we cannot be sure.

@@

The chronological sequence of the cemeteries has received much attention. It has been claimed that no tomb crosses the LM IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), but there is a strong possibility that this is a classificatory artefact [@Hallager2010]. The earliest tomb appears to be the T. 200+ complex, with its precocious cremations and imported artefacts. There remain ongoing debates about whether the KNC represents a foundation on virgin soil, as instances of BA tombs reused the EIA are known (e.g. the Khaniale Teke tholos). Cavanagh [-@Cavanagh1996], for one, could not rule out the possibility of some KNC tombs being of BA date based on his cluster analysis, but other evidence would suggest this is unlikely.

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles, and patterns of tomb construction and abandonment. In the Protogeometric period, some 22 tombs then in use in the KNC, and nine at Fortetsa, ceased to receive interments, followed in the succeeding Protogeometric B period by the founding of 22 new tombs (16 and six respectively). The PGB was defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”. Apparently a Cretan anomaly, this phase sees Attic geometric motifs mixed with experimental patterns possibly deriving from Near Eastern metalwork, as well as iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4). Coldstream has argued that the concurrence of this style and a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T. 107; see @Coldstream1984b; -@Coldstream1994, 112–3). Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-5], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.

The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of new burials around 630 BCE. There had been changes in tomb use in both cemeteries in the preceding century, with no new tombs built in Fortetsa, and few in the KNC, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3]. This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap'. The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown. Although various explanations have been forwarded, [@Coldstream1999, 301-2; @Huxley1994, 126; @Kotsonas2002, 41-4], it seems there are now strong reasons to doubt the settlement was truly abandoned (Whitelaw, pers comm.).

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai [-@Appadurai1986] and Kopytoff [-@Kopytoff1986]. Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence. In Aegean archaeology, these ideas proved attractive because of their resonance with the Homeric motif of gift-giving accompanied by accounts of an object's illustrious heritage [@Whitley2002, 220-1]. It is hardly suprising that, when direct parallels for such objects have appeared archaeologically – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) – a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.

Catling’s discussion of the T. 200+ complex is a notable example. Tomb 201 contained the remains of at least two adults, and possibly a child. The burial objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned BA-style boar’s tusk helmet [@Catling1995, 123]. Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi. All contained imported items at an early date, were of ‘warrior grave’ type, and formed loci for later burials. Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing with precious artefacts and stories of distant lands – ‘heroes returned’ in Catling’s phrase – would have been accorded exceptional treatment in death [@Catling1995], 127–8].

Complementing the theme of heroes abroad has been a similar degree of interest in foreigners at home. Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a known Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician, practising his trade at Knossos. Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9]. While certainty in such individual cases may be moot, it seems unlikely that traders or craftsmen of non-Cretan heritage never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].

Finally, a recurrent interest has been in the 'presence of the past' in these tombs. The practice of hero or ancestor cult, much discussed on the mainland (see @Antonaccio1994; @Coldstream1976), is not common here. Instead, we see several Late Minoan larnakes (and a few other vessels) redeposited in tombs and pits in the KNC and Fortetsa from the PGB period onward. A thorough analysis of these artefacts has been provided by Crowe (2016), who argues that the larnakes were robbed from tombs in the Knossos area, perhaps with a particular preference for painted specimens where available. A particularly evocative case comres from tomb 107, where parts of an elaborate figured LM larnax have been found along with a PGB vessel by the so-called Tree Painter. Both feature women in the company of birds and spiralling trees, and it seems certain that the former inspired the design of the latter. The PGB vessel also seems to draw on Oriental motifs, with the polos crown worn by the goddess having Near Eastern origins (Coldstream 1984a, 94–101). The magpie-like amalgamation of these styles and motifs seem to exemplify the particular cultural openness, fluidity and competition suggested to characterise the PGB period. Artefacts such as this hint towards an intimate discourse of (re)discovery; a painter, struck by the strangeness, or perhaps the uncanny familiarity, of centuries-old images, incorporated them into his own work.
